MyPDDL: Tools for efficiently creating PDDL domains and problems

### 2019-07-19 paper.tex
Sent to Mauro version

\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{img/}}
\usepackage{tabularx}
\newcommand{\tableheadline}[1]{\multicolumn{1}{c}{#1}}
\usepackage{dirtree}
\usepackage{rotating}
\usepackage{caption}
\usepackage{booktabs}
% More space between rows:		
\renewcommand{\arraystretch}{1.1}		
% Slightly more space between columns:		
\setlength{\tabcolsep}{8pt}
\usepackage{minted}
\definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
\DeclareUnicodeCharacter{2212}{-}

@@

\newcommand{\mypddl}{\textsc{myPddl}\xspace}
\newcommand{\mypddlclojure}{\textsc{myPddl-clojure}\xspace}
\newcommand{\mypddlsyntax}{\textsc{myPddl-syntax}\xspace}
\newcommand{\mypddldiagram}{\textsc{myPddl-diagram}\xspace}
\newcommand{\mypddlnew}{\textsc{myPddl-new}\xspace}
\newcommand{\mypddlide}{\textsc{myPddl-ide}\xspace}
\newcommand{\mypddlsnippet}{\textsc{myPddl-snippet}\xspace}
\newcommand{\mypddldistance}{\textsc{myPddl-distance}\xspace}
\newcommand{\mypddlplan}{\textsc{myPddl-plan}\xspace}
\newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
\newcommand{\epd}{editor.planning.domains\xspace}
\newcommand{\itsimple}{\textsc{itSimple}\xspace}
\newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
\newcommand{\pddl}{\textsc{pddl}\xspace}
\newcommand{\uml}{\textsc{uml}\xspace}
\newcommand{\ide}{\textsc{ide}\xspace}
\newcommand{\sublimetext}{Sublime Text\xspace}

@@

Being a key aspect of artificial intelligence, \emph{planning} is
concerned with devising a sequence of actions to achieve a desired
goal \cite{helmert2008understanding}. It can be both a tool to create
automated systems and a means to support and understand human behavior
\cite{konar1999artificial}. However, the effectiveness of planning
largely depends on the quality of the problem formalization
\cite{shah2013knowledge,keps2014}. To ensure a standardized modeling
format, \pddl (\emph{Planning Domain Definition Language})
\cite{mcdermott1998pddl} was developed and has become the de facto
standard for the description of planning tasks
\cite{ilghami2005extension}. The discipline that deals with the
integration of world information into a computer system via a human
expert is called knowledge engineering \cite{feigenbaum1983fifth}.
While automated planning could save vast amounts of time if everything
works as intended, creating the planning task specifications is a
complex task that can be error-prone and cumbersome.

In this chapter, we describe \mypddl
(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
that supports the knowledge engineer in the entire design cycle of
specifying planning tasks. In the initial stages, it allows for the
creation of structured \pddl projects that should encourage a
disciplined design process. With the help of snippets, that is, code
templates, often used syntactic constructs can be inserted into \pddl
files. A syntax highlighting feature that speeds up the
error-detection can come in handy in intermediate
stages. Understanding the textual representation of complex type
hierarchies in domain files can be confusing, so an additional tool
enables their visualization. \pddl's limited modeling capabilities
were bypassed by developing an interface that converts \pddl code into
code of the functional programming language Clojure
\cite{hickey2008clojure} and vice versa.  Within this project, the
interface was employed for a feature that calculates distances between
objects specified in a problem model, but the interface provides
numerous other possibilities and could also be used to further
automate the modeling process. All of the features were integrated
into the customizable and extensible Sublime Text \cite{sublimetext3}
editor. Since the main aim in the development of the toolkit was for
it to be easy to use and maintain, it is evaluated with regard to
these criteria. The usability was assessed by means of a user test
with eight subjects that had no prior experience with AI planning. The
results indicate that both error-detection and the understanding of a
given domain can be facilitated by \mypddl. This chapter is an
extended version of work published in a previous paper~\cite{StrKir2014:aai}.

\begin{figure}[H]
  \centering
  \tikzstyle{every node}=[draw=black,anchor=west,
  align=left,
  scale=0.7]
  \begin{tikzpicture}[%
  grow via three points={one child at (0.5,-0.7) and
    two children at (0.5,-0.7) and (0.5,-1.4)},
  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
  \node {~~~\textbf{myPDDL}~~~}
    child { node {\textbf{-new}\\Create new \pddl project}}		
    child { node {\textbf{-syntax}\\Context-aware syntax highlighting}}
    child { node {\textbf{-snippet}\\Code templates}}
    child { node {\textbf{-clojure}\\Preprocessor for \pddl files via Clojure}}
    child { node {\textbf{-distance}\\Distance calculation in spatial domains}}
    child { node {\textbf{-diagram}\\Visualization of type hierarchy}}
    child { node {\textbf{-plan}\\Planner integration}};
  \end{tikzpicture}
    \caption{\mypddl is a highly customizable and extensible modular
    system, designed for supporting knowledge engineers in the process
    of writing, analyzing and expanding PDDL files and thereby
    promoting the collaboration between knowledge engineers and the
    use of PDDL in real-world applications. It consists of the parts
    shown in the figure.}
  \label{fig:mypddl-overview}  
\end{figure}

The remainder of this chapter is structured as
follows. Section~\ref{sec:related-work} compares \pddl knowledge
engineering tools to lay a foundation for
\mypddl. Section~\ref{sec:mypddl} describes the different modules of
\mypddl and their design principles. Section~\ref{sec:valid-eval}
evaluates \mypddl via a user test. Section~\ref{sec:conclusion}
concludes the chapter and outlines future work.

@@

\pddlstudio \cite{chomuttool,plch2012inspect} is an IDE
(\emph{integrated development environment}) for creating and managing
\textsc{pddl} projects, that is, a collection of \textsc{pddl}
files. Its main features are syntax highlighting, error detection,
context sensitive code completion, code folding, project management,
and planner integration. Many of these features are based on a parser,
which continuously analyzes the code and divides it into syntactic
elements. These elements and the way in which they relate to each
other can then be identified. The syntax highlighter is a tool that
colors constructs according to their syntactical meaning within the
code. In the case of \pddlstudio, it colors names, variables, errors,
keywords, predicates, types, and brackets each in a different
customizable color.  \pddlstudio's error detection can recognize both
syntactic (missing keywords, parentheses, etc.) and semantic (wrong
type of predicate parameters, misspelled predicates, etc.)
errors. This means that \pddlstudio can detect errors due to a
mismatch between domain and problem file in real time. The code
completion feature offers recommendations for standard standard
\textsc{pddl} constructs as well as for previously used terms. Code
folding allows the knowledge engineer to hide currently not needed
code blocks. In this case only the first line of the block is
displayed. All these above mentioned features of \pddlstudio utilize
the parser. To keep track of all files, to display them in a tree
structure, and to save them upon compilation, a project manager is
integrated. Furthermore, this feature is necessary to detect interfile
errors and to use the code completion functionality. Lastly, a
command-line interface allows the integration of planners in order to
run and compare different planning software.

Unlike \pddlstudio, which provides a text based editor for \pddl, the
\itsimple \cite{vaquero2005itsimple} editor has, as its main feature,
a graphical approach that allows designing planning tasks in an
object-oriented approach using \uml (\emph{Unified Modeling
  Language}). In the process leading up to \itsimple,
\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed, a
\textsc{uml} variant specifically designed for modeling planning
domains and problems \cite{vaquero2006use}.

The main purpose of \itsimple is supporting knowledge
engineers in the initial stages of the design phase by making tools
available that help transform the informality of the real world to
formal specifications of domain models. The professed aim of the
project is to provide a means to a \enquote{disciplined process of
  elicitation, organization and analysis of requirements}
\cite{vaquero2005itsimple}. However, subsequent design stages are also
supported. Once domain and problem models have been created,
\textsc{pddl} representations can be generated from the \textsc{uml.p}
diagrams, edited, and then used as input to a number of different
integrated planning systems. Compared to \pddlstudio, \itsimple has a
different approach to planner integration: domain and problem can be
fed to the planner by pressing a button, whereas in \pddlstudio, the
user has to know and input commands in a command-line interface.

With \itsimple, it is possible to directly input the domains and
problems into a planner and to inspect the output from the planning
system using the built-in plan analysis. This consists of a plan
visualization, which shows the interaction between the plan and the
domain by highlighting every change caused by an action. \itsimple's
modeling workflow is unidirectional as changes in the \pddl domain do
not affect the \uml model and \uml models have to be modeled manually,
meaning that they cannot by generated using \pddl. Starting in version
4.0 \cite{vaquero2012itsimple4} \itsimple expanded its features to
allow the creation of \pddl projects from scratch (i.e. without the
\uml to \pddl translation process). Thus far, the \pddl editing
features are basic. A minimal syntax highlighting feature recognizes
\pddl keywords, variables, and comments. Furthermore, \itsimple
provides templates for \pddl constructs, such as requirement
specifications, predicates, actions, initial, and goal definitions.

@@

The online tool \epd allows for editing \pddl files in a web
browser. Its feature comprise syntax highlighting, code folding,
\pddl-specific auto-completion, and multi-tab support. The editor is
part of the Planning.Domains initiative which aims at providing three
pillars to the planning community: (1) an API to access existing \pddl
domains and problems; (2) a planner-in-the-cloud service; and (3) an
online PDDL editor. The online editor is also connected to the
planner-in-the-cloud.

\epd's syntax highlighting features are basic; so far, it does not
support context-aware syntax highlighting: for example,
editor.planning.domains highlights \pddl key words even if they are
misplaced. In contrast, \mypddl is aware of the scope of a construct
and highlights key words accordingly.

The PDDL plugin
vscode-\pddl\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
for the editor Visual Studio Code offers a wide range of editing
functions, such as syntax highlighting, code completion, and code
snippets. It offers a mature planner integration and plan
visualization. Thanks to a \pddl parser integration, it is possible to
detect semantic errors immediately when they are made.

@@

\pddlstudio, \itsimple, and \epd for the most part do not build on
existing editors and therefore cannot fall back on refined
implementations of features, such as selection of tab size, defining
custom key shortcuts, customizing the general look and feel, and
bracket matching. In contrast, vscode-\pddl is integrated into the
mature editor Visual Studio Code and can be used in combination with
other plugins. To have both basic editor features and a high
customizability, it was decided to use an existing, extensible text
editor to integrate \mypddl into.

The tools can also be compared in terms of their syntax highlighting
capabilities. In \pddlmode for Emacs (up to \pddl\,2.2) and \epd (up
to \pddl\,3.1), and vscode-\pddl (up to \pddl\,3.1) keywords,
variables, and comments are highlighted. However, this is only done
via pattern matching without controlling for context. This means that
wherever the respective terms appear within the code they will get
highlighted, regardless of the syntactical correctness. Therefore, it
is useful when the knowledge engineer is familiar with \pddl syntax
but can also be misleading if this is not the case. Different colors
can be chosen by customizing Emacs and Visual Studio Code. \epd
provides two fixed color schemes. \itsimple's syntax highlighting for
\pddl 3.1 is, except for the \pddl version difference, equally as
extensive as that of \pddlmode for Emacs but does not allow for any
customization.  \pddlstudio has advanced syntax highlighting that
distinguishes all different \pddl 1.2 constructs depending on the
context and allows knowledge engineers to choose their preferred
highlighting colors. One of the primary objectives of \mypddl is to
help users in keeping track of their \pddl programs. As a means to
this end, it was decided to also implement sophisticated,
context-dependent syntax highlighting.

Another feature that can be useful for fast programming is the ability
to insert larger code skeletons or snippets. This allows the knowledge
engineer to focus on the specific domain and problem characteristics
instead of having to worry about the \pddl formalities. \pddlstudio
does not support the insertion of code snippets. \itsimple features
some code templates for predicates, derived predicates, functions,
actions, constraints, types, comments, requirements, objects, and
metrics. However, the templates are neither customizable nor
extensible. \pddlmode for Emacs provides three larger skeletons: one
for domains, one for problems, and one for actions. Further skeletons
could be added. Both \epd and vscode-\pddl provide many code
snippets. \mypddl aims to combine the best of these latter tools and
support customizable and extensible snippets for domains, problems,
types, predicates, functions, actions, and durative actions.

\pddlstudio, \pddlmode for Emacs, and \epd do not provide
visualization options. \itsimple, on the other hand, is based entirely
on visually modeling domains and problems. Therefore, since the first
version, the focus has mainly been on exporting from \textsc{uml.p} to
\pddl. \mypddl is to reverse this design approach and enable type
diagram visualization of some parts of the \pddl code. vscode-\pddl
does not provide domain visualizationa but is able to visualize a
found plan.

Searching for errors can be one of the most time consuming parts of
the design process. Hence, any tool that is able to help detect errors
faster is of great value to the knowledge engineer. While \pddlmode
for Emacs, \itsimple, and \epd facilitate error detection only by
basic syntax highlighting, both \pddlstudio and vscode\-pddl are able
to detect errors via a \pddl parser. In \mypddl, syntactic errors are
not be highlighted by the syntax highlighting feature, while all
correct \pddl code is to get highlighted.

A major drawback of \pddlstudio and \pddlmode for Emacs especially is
that they are not updated regularly to support the most recent \pddl
versions. \pddlstudio's parser is only able to parse \pddl 1.2
\cite{chomuttool}, while the latest \pddl version is 3.1. \pddl has
significantly evolved since \pddl 1.2 and was extended in \pddl 2.1 to
include \emph{durative actions} to model time dependent behaviors,
\emph{numeric fluents} to model non-binary changes of the world state,
and \emph{plan-metrics} to customize the evaluation of plans
\cite{fox2003pddl2}. \pddlmode for Emacs is only compatible with \pddl
versions up to 2.2, which introduced \emph{derived predicates} and
\emph{timed initial predicates} \cite{edelkamp2004pddl2} but does not
recognize later features like \emph{object-fluents}. It follows that
the range of functions specified in the domain file cannot include
object-types in addition to numbers. \itsimple, \epd, vscode-\pddl,
and \mypddl support the latest \pddl version.

An important features of any software is the possibility of extending
and customizing it \cite{hui2003requirements}. \pddlstudio falls short
of satisfying this requirement as the customization options are
limited to the choice of font style and color of highlighted \pddl
expressions. Furthermore, \pddlstudio is written as standalone
program, meaning that there are no \pddl independent extensions. The
same holds true for \itsimple. Since both Emacs and VS Code are
established editors, \pddlmode and vscode-\pddl are highly and easily
customizable and extensible. This is the other major reason why it was
decided that \mypddl should be integrated into a existing, extensible,
and customizable text editor. These requirements are intended to be
met by Sublime Text, a text editor that offers such features as
customizable key bindings, display of line numbers, and multi-line
selection.

@@

  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{planning.domains} & \tableheadline{VS Code} & \tableheadline{\mypddl} \\

@@

  \uml to \pddl translation           & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & yes       \\
  planner integration                 & allowing for convenient planner access                & basic                       & yes                       & no                            &   yes    & yes     & yes       \\

@@

\begin{description}
  \setlength\itemsep{1em}
\item[myPDDL-IDE] is an integrated development environment (IDE) for
  the use of \mypddl in the text and code editor \emph{Sublime
    Text}\footnote{\url{http://www.sublimetext.com}}. Since
  \mypddlsnippet and \textsc{-syntax} are devised explicitly for
  \sublimetext, their integration is implicit. The other tools
  described below (\mypddl-new, -diagram, -distance, -plan) can be
  used independently of \sublimetext via the command-line but can also
  be called from the editor.

\item[myPDDL-syntax] is a context-aware syntax highlighting feature
  for \sublimetext. It recognizes all \pddl constructs up to version
  3.1, such as comments, variables, names, and keywords and highlights
  them in different colors. Using regular expressions and a
  sophisticated pattern matching heuristic, it detects both the start
  and the end of \pddl code blocks and constructs. It then divides
  them into \emph{scopes}, that is, named regions. \sublimetext
  colorizes the code elements via the assigned scope names and in
  accordance with the current color scheme. These scopes allow for a
  fragmentation of the \pddl files, so that constructs are only
  highlighted if they appear in the correct context. Thus missing
  brackets, misplaced expressions and misspelled keywords are visually
  distinct and can be identified (see Figure~\ref{fig:syntax}).

\item[myPDDL-new] helps to organize \pddl projects. In many cases
  \pddl domains are created ad hoc \cite{shah2013exploring}. However,
  each implementation of a \pddl task specification comprises one
  domain and at least one corresponding problem file. Since several
  team members may be working on these files, keeping \pddl projects
  organized will facilitate collaboration. An automatically created,
  standardized project folder structure could facilitate the
  collaboration between users and the maintenance of consistency
  across projects. To this end, \mypddlnew creates the following
  folder structure when creating a new \pddl project:
  \begin{figure}[H]
  \dirtree{%
  .1 project-name/.
  .2 domains/.
  .2 problems/.
  .3 p01.pddl.
  .2 solutions/.
  .2 domain.pddl.
  .2 README.md.
  .2 plan.
  }
\end{figure}
  All of the templates to create the files can be customized and new
  templates can be added. The domain file \texttt{domain.pddl} and the
  problem file \texttt{p01.pddl} initially contain corresponding \pddl
  skeletons. Additionally the project name is used as the domain name
  within the files \texttt{domain.pddl} and \texttt{p01.pddl}. All
  problem files that are associated with one domain file are collected
  in the folder \texttt{problems/}. \texttt{README.md} is a Markdown
  file, which is intended for information about the authors of the
  project, contact information, informal domain and problem
  specifications, and licensing information.  Markdown files can be
  converted to \textsc{html} by various hosting services (like GitHub
  or Bitbucket). The basic planner integration \mypddlplan provided by
  the file \texttt{plan} is described below.

@@

For example, typing \texttt{action} and pressing the tabulator key
inserts the action skeleton in Listing \ref{ls:action-skeleton}. \pddl
constructs with a specified arity can be generated by adding the arity
number to the trigger (\texttt{p2} would insert the binary predicate
template \texttt{(pred-name~?x~-~object~?y~-~object)}.

Every snippet is stored in a separate file, located in the folder
\texttt{Packages/PDDL/} of Sublime Text. New snippets can be added and
existing snippets can be customized by changing the templates in this
folder.

>>>>

\usepackage{hyperref}
\usepackage{graphicx}
\graphicspath{{img/}}
\usepackage{tabularx}
\newcommand{\tableheadline}[1]{\multicolumn{1}{c}{#1}}
\usepackage{dirtree}
\usepackage{rotating}
\usepackage{caption}
\usepackage{booktabs}

% More space between rows:		
\renewcommand{\arraystretch}{1.1}		
% Slightly more space between columns:		
\setlength{\tabcolsep}{8pt}
\usepackage{minted}
\definecolor{dhscodebg}{rgb}{0.95,0.95,0.95} % listing background
%\DeclareUnicodeCharacter{2212}{-}

@@

\newcommand{\mypddl}{\textsc{myPddl}\xspace}
\newcommand{\mypddlclojure}{\textsc{myPddl-clojure}\xspace}
\newcommand{\mypddlsyntax}{\textsc{myPddl-syntax}\xspace}
\newcommand{\mypddldiagram}{\textsc{myPddl-diagram}\xspace}
\newcommand{\mypddlnew}{\textsc{myPddl-new}\xspace}
\newcommand{\mypddlide}{\textsc{myPddl-ide}\xspace}
\newcommand{\mypddlsnippet}{\textsc{myPddl-snippet}\xspace}
\newcommand{\mypddldistance}{\textsc{myPddl-distance}\xspace}
\newcommand{\mypddlplan}{\textsc{myPddl-plan}\xspace}
\newcommand{\pddlstudio}{\textsc{pddl studio}\xspace}
\newcommand{\epd}{editor.planning.domains\xspace}
\newcommand{\itsimple}{\textsc{itSimple}\xspace}
\newcommand{\pddlmode}{\textsc{pddl}-mode\xspace}
\newcommand{\vscode}{vscode-\textsc{pddl}\xspace}
\newcommand{\pddl}{\textsc{pddl}\xspace}
\newcommand{\uml}{\textsc{uml}\xspace}
\newcommand{\ide}{\textsc{ide}\xspace}
\newcommand{\sublimetext}{Sublime Text\xspace}

@@

Being a key aspect of artificial intelligence, \emph{planning} is
concerned with devising a sequence of actions to achieve a desired
goal \cite{helmert2008understanding}. It can be both a tool to create
automated systems and a means to support and understand human
behavior~\cite{konar1999artificial}. However, the effectiveness of
planning largely depends on the quality of the problem formalization
\cite{shah2013knowledge}. To ensure a standardized modeling format,
\pddl (\emph{Planning Domain Definition Language})
\cite{mcdermott1998pddl} was developed and has become the de facto
standard for the description of planning tasks
\cite{ilghami2005extension}. The discipline that deals with the
integration of world information into a computer system via a human
expert is called knowledge engineering \cite{feigenbaum1983fifth}.
While automated planning could save vast amounts of time if everything
works as intended, creating the planning task specifications is a
complex task that can be error-prone and cumbersome.

In this chapter, we describe \mypddl
(Figure~\ref{fig:mypddl-overview}), a knowledge engineering toolkit
that supports the knowledge engineer in the entire design cycle of
specifying planning tasks. In the initial stages, it allows for the
creation of structured \pddl projects that should encourage a
disciplined design process. With the help of snippets, that is, code
templates, often used syntactic constructs can be inserted into \pddl
files. A syntax highlighting feature that speeds up the
error-detection can come in handy in intermediate
stages. Understanding the textual representation of complex type
hierarchies in domain files can be confusing, so an additional tool
enables their visualization. \pddl's limited modeling capabilities
were bypassed by developing an interface that converts \pddl code into
code of the functional programming language Clojure
\cite{hickey2008clojure} and vice versa.  Within this project, the
interface was employed for a feature that calculates distances between
objects specified in a problem model, but the interface provides
numerous other possibilities and could also be used to further
automate the modeling process. A basic planner integration allows for
quickly running a desired planner. All of the features were integrated
into the customizable and extensible Sublime
Text\footnote{\url{http://www.sublimetext.com/}} editor. Since the
main aim in the development of the toolkit was for it to be easy to
use and maintain, it is evaluated with regard to these criteria. The
usability was assessed by means of a user test with eight subjects
that had no prior experience with AI planning. The results indicate
that both error-detection and the understanding of a given domain can
be facilitated by \mypddl.

\begin{figure}[H]
  \centering
  \tikzstyle{every node}=[draw=black,anchor=west,
  align=left,
  scale=0.7]
  \begin{tikzpicture}[%
  grow via three points={one child at (0.5,-0.7) and
    two children at (0.5,-0.7) and (0.5,-1.4)},
  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
  \node {~~~\textbf{myPDDL}~~~}
  child { node {\textbf{-new}\\Create new \pddl project}}
  child { node {\textbf{-snippet}\\Code templates}}
  child { node {\textbf{-syntax}\\Context-aware syntax highlighting}}
  child { node {\textbf{-clojure}\\Preprocessor for \pddl files via Clojure}}
  child { node {\textbf{-distance}\\Distance calculation in spatial domains}}
  child { node {\textbf{-diagram}\\Visualization of type hierarchy}}
  child { node {\textbf{-plan}\\Planner integration}};
  \end{tikzpicture}
    \caption{\mypddl is a highly customizable and extensible modular
    system, designed for supporting knowledge engineers in the process
    of writing, analyzing and expanding PDDL files and thereby
    promoting the collaboration between knowledge engineers and the
    use of PDDL in real-world applications. It consists of the parts
    shown in the figure.}
  \label{fig:mypddl-overview}  
\end{figure}

This chapter is an extended version of work published in a previous
paper~\cite{StrKir2014:aai}. The remainder of this chapter is
structured as follows. Section~\ref{sec:related-work} compares \pddl
knowledge engineering tools to lay a foundation for
\mypddl. Section~\ref{sec:mypddl} describes the different modules of
\mypddl and their design principles. Section~\ref{sec:valid-eval}
evaluates \mypddl via a user test. Section~\ref{sec:conclusion}
concludes the chapter and outlines future work.

@@

\pddlstudio \cite{plch2012inspect} is an IDE (\emph{integrated
  development environment}) for creating and managing \textsc{pddl}
projects, that is, a collection of \textsc{pddl} files. Its main
features are syntax highlighting, error detection, context sensitive
code completion, code folding, project management, and planner
integration. Many of these features are based on a parser, which
continuously analyzes the code and divides it into syntactic
elements. These elements and the way in which they relate to each
other can then be identified. The syntax highlighter is a tool that
colors constructs according to their syntactical meaning within the
code. In the case of \pddlstudio, it colors names, variables, errors,
keywords, predicates, types, and brackets each in a different
customizable color.  \pddlstudio's error detection can recognize both
syntactic (missing keywords, parentheses, etc.) and semantic (wrong
type of predicate parameters, misspelled predicates, etc.)
errors. This means that \pddlstudio can detect errors based on a
mismatch between domain and problem file in real time. The code
completion feature offers recommendations for standard standard
\textsc{pddl} constructs as well as for previously used terms. Code
folding allows the knowledge engineer to hide currently not needed
code blocks. In this case only the first line of the block is
displayed. Lastly, a command-line interface allows the integration of
planners in order to run and compare different planning software.

Unlike \pddlstudio, which provides a text based editor for \pddl, the
\itsimple \cite{vaquero2005itsimple} editor has, as its main feature,
a graphical approach that allows designing planning tasks in an
object-oriented approach using \uml (\emph{Unified Modeling
  Language}). In the process leading up to \itsimple,
\textsc{uml.p}---\textsc{uml} in a Planning Approach---was proposed,
which is a \textsc{uml} variant specifically designed for modeling
planning domains and problems \cite{vaquero2006use}.

The main purpose of \itsimple is supporting knowledge engineers in the
initial stages of the design phase by making tools available that help
transform the informality of the real world to formal specifications
of domain models. The professed aim of the project is to provide a
means to a \enquote{disciplined process of elicitation, organization
  and analysis of requirements} \cite{vaquero2005itsimple}. However,
subsequent design stages are also supported. Once domain and problem
models have been created, \textsc{pddl} representations can be
generated from the \textsc{uml.p} diagrams, edited, and then used as
input to a number of different integrated planning systems.

With \itsimple, it is possible to directly input the domains and
problems into a planner and to inspect the output from the planning
system using the built-in plan analysis. This consists of a plan
visualization, which shows the interaction between the plan and the
domain by highlighting every change caused by an action. \itsimple's
modeling workflow is unidirectional as changes in the \pddl domain do
not affect the \uml model and \uml models have to be modeled manually,
meaning that they cannot by generated using \pddl.

@@

The online tool \epd allows for editing \pddl files in a web
browser. Its feature comprise syntax highlighting, code folding,
\pddl-specific auto-completion, and multi-tab support. The editor is
part of the Planning-Domains\footnote{\url{http://planning.domains/}}
initiative which aims at providing three pillars to the planning
community: (1) an API to access existing \pddl domains and problems;
(2) a planner-in-the-cloud service which be be accessed via a RESTful
API; and (3) an online \pddl editor. The online editor is also
connected to the planner in the cloud.

The \pddl plugin
\vscode\footnote{\url{https://marketplace.visualstudio.com/items?itemName=jan-dolejsi.pddl}}
for the editor VS Code (\emph{Visual Studio Code}) offers a wide range
of editing functions, such as syntax highlighting, code completion,
code folding, and code snippets. It offers a mature planner
integration and plan visualization. Thanks to a \pddl parser
integration, it is possible to detect semantic errors immediately when
they are made.

@@

\pddlstudio, \itsimple, and \epd for the most part do not build on
existing editors and therefore cannot fall back on refined
implementations of features, such as selection of tab size, defining
custom key shortcuts, customizing the general look and feel, and
bracket matching. In contrast, \vscode is integrated into the
mature editor Visual Studio Code and can be used in combination with
other plugins. To have both basic editor features and a high
customizability, it was decided to use an existing, extensible text
editor to integrate \mypddl into.

The tools can also be compared in terms of their syntax highlighting
capabilities. In \pddlmode for Emacs (up to \pddl\,2.2) and \epd (up
to \pddl\,3.1), and \vscode (up to \pddl\,3.1) keywords, variables,
and comments are highlighted. However, this is only done via pattern
matching without controlling for context. This means that wherever the
respective terms appear within the code they will get highlighted,
regardless of the syntactical correctness. Different colors can be
chosen by customizing Emacs and Visual Studio Code. \epd provides two
fixed color schemes. \itsimple's syntax highlighting for \pddl 3.1 is,
except for the \pddl version difference, equally as extensive as that
of \pddlmode for Emacs but does not allow for any customization.
\pddlstudio has advanced syntax highlighting that distinguishes all
different \pddl 1.2 constructs depending on the context and allows
knowledge engineers to choose their preferred highlighting colors. One
of the primary objectives of \mypddl is to help users in keeping track
of their \pddl programs. As a means to this end, it was decided to
also implement sophisticated, context-dependent syntax highlighting.

Another feature that can be useful for fast development is the ability
to insert larger code skeletons or snippets. \pddlstudio does not
support the insertion of code snippets. \itsimple features some code
templates for predicates, derived predicates, functions, actions,
constraints, types, comments, requirements, objects, and
metrics. However, the templates are neither customizable nor
extensible. \pddlmode for Emacs provides three larger skeletons: one
for domains, one for problems, and one for actions. Further skeletons
could be added. Both \epd and \vscode provide many code
snippets. \mypddl aims to combine the best of these latter tools and
support customizable and extensible snippets for domains, problems,
types, predicates, functions, actions, and durative actions.

\pddlstudio, \pddlmode for Emacs, and \epd do not provide
visualization options. \itsimple, on the other hand, is based entirely
on visually modeling domains and problems. Therefore, since the first
version, the focus has mainly been on exporting from \textsc{uml.p} to
\pddl and to visualize plans. \mypddl is to reverse this design
approach and enable type diagram visualization of some parts of the
\pddl code. \vscode does not provide domain visualization but is able
to visualize a found plan.

Searching for errors can be one of the most time consuming parts of
the design process. Hence, any tool that is able to help detect errors
faster is of great value to the knowledge engineer. While \pddlmode
for Emacs, \itsimple, and \epd facilitate error detection only by
basic syntax highlighting, both \pddlstudio and \vscode are able to
detect errors via a \pddl parser. In \mypddl, a different approach is
taken and syntactic errors are \emph{not} highlighted by the syntax
highlighting feature, while all correct \pddl code \emph{is}
highlighted.

A major drawback of \pddlstudio and \pddlmode for Emacs especially is
that they are not updated regularly to support the most recent \pddl
versions. \pddlstudio's parser is only able to parse \pddl 1.2, while
the latest \pddl version is 3.1. \pddl has significantly evolved since
\pddl 1.2 and was extended in \pddl 2.1 to include \emph{durative
  actions} to model time dependent behaviors, \emph{numeric fluents}
to model non-binary changes of the world state, and
\emph{plan-metrics} to customize the evaluation of plans. \pddlmode
for Emacs is only compatible with \pddl versions up to 2.2, which
introduced \emph{derived predicates} and \emph{timed initial
  predicates} but does not recognize later features like
\emph{object-fluents}. It follows that the range of functions
specified in the domain file cannot include object-types in addition
to numbers. \itsimple, \epd, \vscode, and \mypddl support the latest
\pddl version.

\pddlstudio falls short of customization options since they are
limited to the choice of font style and color of highlighted \pddl
expressions. Furthermore, \pddlstudio is written as standalone
program, meaning that there are no \pddl-independent extensions. The
same holds true for \itsimple. Since both Emacs and VS Code are
established editors, \pddlmode and \vscode are highly customizable and
extensible. This is the other major reason why it was decided that
\mypddl should be integrated into a existing, extensible, and
customizable text editor. These requirements are met by Sublime Text,
a text editor that offers a wide variety of features and plugins.

@@

  \tableheadline{Feature}             & \tableheadline{Function}                              & \tableheadline{\pddlstudio} & \tableheadline{\itsimple} & \tableheadline{\pddlmode} & \tableheadline{planning.domains} & \tableheadline{\vscode} & \tableheadline{\mypddl} \\

@@

  \uml to \pddl translation           & supporting initial modeling                           & no                          & yes                       & no                            &   no     & no      & no       \\
  planner integration                 & allowing for convenient planner access                & yes                       & yes                       & no                            &   yes    & yes     & yes       \\

@@

\begin{description}
  \setlength\itemsep{1em}
\item[myPDDL-IDE] is an integrated development environment (IDE) for
  the use of \mypddl in the text and code editor \emph{Sublime
    Text}\footnote{\url{http://www.sublimetext.com}}. Since
  \mypddlsnippet and \textsc{-syntax} (see below) are devised
  explicitly for \sublimetext, their integration is implicit. The
  other tools described below (\mypddl-new, -diagram, -distance,
  -plan) can be used independently of \sublimetext via the
  command-line but can also be called from the editor.

\item[myPDDL-new] helps to organize \pddl projects. In many cases
  \pddl domains are created ad hoc \cite{shah2013exploring}. However,
  each implementation of a \pddl task specification comprises one
  domain and at least one corresponding problem file. Since several
  team members may be working on these files, keeping \pddl projects
  organized will facilitate collaboration. An automatically created,
  standardized project folder structure could facilitate the
  collaboration between users and the maintenance of consistency
  across projects. To this end, \mypddlnew creates the following
  folder structure when creating a new \pddl project:
  \begin{figure}[H]
  \dirtree{%
  .1 project-name/.
  .2 domains/.
  .2 problems/.
  .3 p01.pddl.
  .2 solutions/.
  .2 domain.pddl.
  .2 README.md.
  .2 plan.
  }
\end{figure}
All of the templates to create the files can be customized and new
templates can be added. The domain file \texttt{domain.pddl} and the
problem file \texttt{p01.pddl} initially contain corresponding \pddl
skeletons. Additionally the project name is used as the domain name
within the files \texttt{domain.pddl} and \texttt{p01.pddl}. All
problem files that are associated with one domain file are collected
in the folder \texttt{problems/}. \texttt{README.md} is a Markdown
file, which is intended for information about the authors of the
project, contact information, informal domain and problem
specifications, and licensing information.  Markdown files can be
converted to \textsc{html} by various hosting services like GitHub or
Bitbucket. The basic planner integration \mypddlplan provided by the
file \texttt{plan} is described below.

@@

For example, typing \texttt{action} and pressing the tabulator key
inserts a skeleton to specify an action. \pddl constructs with a
specified arity can be generated by adding the arity number to the
trigger (\texttt{p2} would insert the binary predicate template
\texttt{(pred-name~?x~-~object~?y~-~object)}.

Every snippet is stored in a separate file, located in the packages
folder of Sublime Text. New snippets can be added and existing
snippets can be customized by changing the templates in this folder.

\item[myPDDL-syntax] is a context-aware syntax highlighting feature
  for \sublimetext. It recognizes all \pddl constructs up to version
  3.1, such as comments, variables, names, and keywords and highlights
  them in different colors. Using regular expressions and a
  sophisticated pattern matching heuristic, it detects both the start
  and the end of \pddl code blocks and constructs. It then divides
  them into \emph{scopes}, that is, named regions. \sublimetext
  colorizes the code elements via the assigned scope names and in
  accordance with the current color scheme. These scopes allow for a
  fragmentation of the \pddl files, so that constructs are only
  highlighted if they appear in the correct context. Thus missing
  brackets, misplaced expressions and misspelled keywords are visually
  distinct and can be identified (Figure~\ref{fig:syntax}).
